---
title: "Assignment 2: Coding Basics"
author: "Rebecca Murphy"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. Creating a sequence of numbers from 1 - 100 by intervals of 4; sequence name is sequence1
seq(1,100,4)
sequence1 <- seq(1,100,4)

#2. Computing the mean and median of sequence1
mean(sequence1)
median(sequence1)

#3. Determining if the mean > median
mean(sequence1) > median(sequence1)

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}

student_names <- c("Sarah", "Chris","Mark", "Rachel") #character vector 
test_scores <- c(100L, 45L, 75L, 88L)  #Integer vector 
pass_or_fail <- c(TRUE, FALSE, TRUE, TRUE) #Logical vector

grading_df <- data.frame(student_names,test_scores,pass_or_fail)
names(grading_df) <- c("Names", "Test Scores", "Pass?")
View(grading_df)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: Although both are 2D structures, by having character, numeric, and logical vectors, the data frame differs from a matrix in the fact that it allows different modes/elements. Matrices would only be able to have one of those vector types.  

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#10: Function that gives "pass" if the grade is above 50 and "fail" if it is below 50 - 
#Note - I didn't need to use print function to get the desired output.
grade_test <- function(x){ 
   ifelse(c(x>50), "Pass", "Fail")
}

#11: applying this function against the test scores from #5
grade_test(test_scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The 'ifelse' statement worked (instead of if and else) because there was more than one component for the test scores vector. If there had only been one test score, I could have used 'if' and 'else'. 
